Untangling the link between traits, size and growth rate in plants

### 2015-02-06 ms.tex
Equations explaining shift in correlation with size. Towards #25.

\subsection{Trait values maximising height
growth}\label{trait-values-maximising-height-growth}

We want to find the trait values maximising growth rate, \(G\). To make
the analysis more tractible, we will focus on height growth rate and
assume we are dealing with a plant of given height where 100\% of
available energy is allocated to growth. From eq. \ref{eq:dhdt}, we thus
have

\begin{equation} \label{eq:G1}
G = \frac{\textrm{d}h}{\textrm{d}t} = \frac{\textrm{d}h}{\textrm{d}a_\textrm{l}}
\times \frac{\textrm{d}a_\textrm{l}}{\textrm{d}m_\textrm{t}}
\times \frac{\textrm{d}m_\textrm{t}} {\textrm{d}P}
\times \frac{\textrm{d}P}{\textrm{d}t}.
\end{equation}

Maximising \(G\) is equivalent to solving for trait values \(x\) giving
\(\partial G /\partial x = 0\). Noting that traits do not influence
\(\frac{\textrm{d}h} {\textrm{d}a_\textrm{l}}\), and by assumption
\(\frac{\textrm{d}m_\textrm{t}}{\textrm{d}P}=1\), we thus have

\begin{equation} \label{eq:G2}
G =  c_1   \left(\frac{\textrm{d}a_\textrm{l}} {\textrm{d}m_\textrm{t}}  \frac{ \textrm{d}P} {\textrm{d}t} \right),
\end{equation}

where \(c_1 = \frac{\textrm{d}h}{\textrm{d}a_\textrm{l}}\). Eq.
\ref{eq:G2} is a product of the form \(Y = W(x) \times Z(x)\). For,
equations of this type, a solution to \(\partial{Y}/\partial{x} =0\) is
given by \(\partial{W}/\partial{x} / W = \partial{Z}/\partial{x} / Z\).
Maximum growth rate thus occurs when

\begin{equation} \label{eq:G3}
\frac{\partial \left(\frac{\textrm{d}a_\textrm{l}} {\textrm{d}m_\textrm{t}}\right)}{\partial x} \frac{1}{\frac{\textrm{d}a_\textrm{l}} {\textrm{d}m_\textrm{t}}} = \frac{\partial \left( \frac{ \textrm{d}P} {\textrm{d}t}\right)}{\partial x} \frac{1}{\frac{ \textrm{d}P} {\textrm{d}t} }.
\end{equation}

In words, the maximum occurs when relative change in marginal leaf
deployment with respect to trait is equal to the relative change in mass
production with respect to the trait.

Let us now try and simplify equation \ref{eq:G3}. First, let our trait
of interest be leaf-construction cost, i.e. \(x=\phi\). From eq/
\ref{eq:daldmt}, we can derive

\begin{equation} \label{eq:G4}
\frac{\partial \left(\frac{\textrm{d}a_\textrm{l}} {\textrm{d}m_\textrm{t}}\right)}{\partial \phi} = -\left(\frac{\textrm{d}a_\textrm{l}} {\textrm{d}m_\textrm{t}}\ \right)^2.
\end{equation}

This simplifies eq. \ref{eq:G3} to

\begin{equation} \label{eq:G5}
\frac{\textrm{d}a_\textrm{l}} {\textrm{d}m_\textrm{t}}(x) =  \frac{\partial \left( \frac{ \textrm{d}P} {\textrm{d}t}\right)}{\partial x} \frac{1}{\frac{ \textrm{d}P} {\textrm{d}t} }.
\end{equation}

From eq. \ref{eq:dPdt} we obtain,

@@

where
\(c_2 = Y ( a_\textrm{l} \, (A - r_\textrm{l}) + \sum_\textrm{i=b,s,r}{m_\textrm{i} \, r_\textrm{i}}) - (\sum_\textrm{i=b,s,r}{m_\textrm{i} \, k_\textrm{i}})\)
is independt of \(\phi\), and thus

\begin{equation}\label{eq:dPdt3}
\frac{\partial \left( \frac{ \textrm{d}P} {\textrm{d}t}\right)}{\partial \phi}  = - (1-B_4) a_\textrm{l} a_4\phi ^{-B_4}.
\end{equation}

Combing eqs. \ref{eq:G5}-\ref{eq:dPdt3}, we find that the maximum occurs
when

\begin{equation} \label{eq:G6}
\frac{\textrm{d}a_\textrm{l}} {\textrm{d}m_\textrm{t}}(\phi) = \frac{- (1-B_4) a_\textrm{l} a_4\phi ^{-B_4}}{c_2 - a_\textrm{l} a_4 \phi ^{1-B_4}}.
\end{equation}

Note also that if \(B4=1\),
\(\frac{\partial \left( \frac{ \textrm{d}P} {\textrm{d}t}\right)}{\partial \phi} =0\),
also if \(B4>1\),
\(\frac{\partial \left( \frac{ \textrm{d}P} {\textrm{d}t}\right)}{\partial \phi} > 0\).
A similar derivation can be produced linking wood density to
productivity, via sapwood turnover.

>>>>

\subsection{Effect of traits on height growth} \label{trait-values-maximising-height-growth}

We want to show how two traits $\phi$ and $\rho$ influence growth rate,
by looking at the effect of
traits on the components of growth rate (eq. \ref{eq:dhdt}. To make
the analysis more tractable, we will assume we are dealing with a plant of
given height where 100\% of available energy is allocated to growth,
i.e. \(\frac{\textrm{d}m_\textrm{t}}{\textrm{d}P}=1\). Note also
that by the assumptions of our model and above, traits do not influence
\(\frac{\textrm{d}h} {\textrm{d}a_\textrm{l}}\). Thus we have

\begin{equation} \label{eq:G2}
G = \frac{\textrm{d}h}{\textrm{d}t} = c_1   \left(\frac{\textrm{d}a_\textrm{l}} {\textrm{d}m_\textrm{t}}  \frac{ \textrm{d}P} {\textrm{d}t} \right),
\end{equation}

where \(c_1 = \frac{\textrm{d}h}{\textrm{d}a_\textrm{l}}\). Eq.
\ref{eq:G2} is a product of the form \(Y(x) = W(x) \times Z(x)\). For,
equations of this type, the derivative \(\partial{Y}/\partial{x}\) is
given by \(\partial{W}/\partial{x} \, Z + W \, \partial{Z}/\partial{x}\). Thus
the derivative of \(G\) with respect to some trait $x$  is given by

\begin{equation} \label{eq:dG}
\frac{\partial G} {\partial x} =
\frac{\partial \left(\frac{\textrm{d}a_\textrm{l}} {\textrm{d}m_\textrm{t}}\right)}{\partial x}
 \, \frac{ \textrm{d}P} {\textrm{d}t}
+ \frac{\textrm{d}a_\textrm{l}} {\textrm{d}m_\textrm{t}}
\, \frac{\partial \left( \frac{ \textrm{d}P} {\textrm{d}t}\right)}{\partial x}.
\end{equation}

If this derivative is positive (negative), there will be a positive (negative) relationship
between traits and growth at the height where it is evaluated.

We further simplify by expanding the derivative of eq. \ref{eq:daldmt}:

\begin{equation} \label{eq:da_dx}
\frac{\partial \left(\frac{\textrm{d}a_\textrm{l}} {\textrm{d}m_\textrm{t}}\right)}
{\partial x} = -\left(\frac{\textrm{d}a_\textrm{l}} {\textrm{d}m_\textrm{t}}\ \right)^2
\, \frac{\partial \left(\frac{\textrm{d}m_\textrm{t}} {\textrm{d}a_\textrm{l}}\right)
}{\partial x}.
\end{equation}

Combing eqs. \ref{eq:dG}-\ref{eq:da_dx}, we have

\begin{equation} \label{eq:dG2}
\frac{\partial G} {\partial x} =
\frac{\textrm{d}a_\textrm{l}} {\textrm{d}m_\textrm{t}}
\left(
\frac{\partial \left( \frac{ \textrm{d}P} {\textrm{d}t}\right)}{\partial x}
- \frac{\textrm{d}a_\textrm{l}} {\textrm{d}m_\textrm{t}}
\,  \frac{\partial \left(\frac{\textrm{d}m_\textrm{t}} {\textrm{d}a_\textrm{l}}\right)
}{\partial x}
 \, \frac{ \textrm{d}P} {\textrm{d}t}
\right).
\end{equation}

Since is $\frac{\textrm{d}a_\textrm{l}} {\textrm{d}m_\textrm{t}}>0$ always, the sign
of $\frac{\partial G} {\partial x}$ depends on the terms inside the brackets on the
RHS. Rearranging we find that $\frac
{\partial G} {\partial x} < 0$ if

\begin{equation}\label{eq:dg3}
\frac{\partial \left( \frac{ \textrm{d}P} {\textrm{d}t}\right)}{\partial x}
< \frac{\textrm{d}a_\textrm{l}} {\textrm{d}m_\textrm{t}}
\,  \frac{\partial \left(\frac{\textrm{d}m_\textrm{t}} {\textrm{d}a_\textrm{l}}\right)
}{\partial x} \, \frac{ \textrm{d}P} {\textrm{d}t},
\end{equation}

or equivalently
\begin{equation}\label{eq:dg4}
\frac{
\frac{\partial \left( \frac{ \textrm{d}P} {\textrm{d}t}\right)}{\partial x} }
{\frac{ \textrm{d}P} {\textrm{d}t}}
<
\frac{ \frac{\partial \left(\frac{\textrm{d}m_\textrm{t}} {\textrm{d}a_\textrm{l}}\right)
}{\partial x}}{\frac{\textrm{d}m_\textrm{t}} {\textrm{d}a_\textrm{l}}}.
\end{equation}

In words, there will be a negative relationship between trait x and growth if the
relative change in production due to change in traits is less than the relative change
in the total construction cost for a unit of leaf area due to change in trait. Crucially,
key elements in eq. \ref{eq:dg3} are modified by height.

We can also solve for trait values \(x\) maximising growth rate, i.e. the value
giving \(\partial G /\partial x = 0\):

\begin{equation}\label{eq:max}
\frac{
\frac{\partial \left( \frac{ \textrm{d}P} {\textrm{d}t}\right)}{\partial x} }
{\frac{ \textrm{d}P} {\textrm{d}t}}
=
\frac{ \frac{\partial \left(\frac{\textrm{d}m_\textrm{t}} {\textrm{d}a_\textrm{l}}\right)
}{\partial x}}{\frac{\textrm{d}m_\textrm{t}} {\textrm{d}a_\textrm{l}}}.
\end{equation}

\subsubsection{Leaf construction cost}

If the trait of interest is leaf-construction cost, i.e. \(x=\phi\), we can expand
eq. \ref{eq:dg4} as follows. Note that $\frac{\partial \left(\frac{\textrm{d}m_\textrm
{t}} {\textrm{d}a_\textrm{l}}\right)}{\partial \phi} = 1$. Also, from eq. \ref{eq:dPdt} we obtain,

@@

where
\(c_2 = Y ( a_\textrm{l} \, (A - r_\textrm{l}) + \sum_\textrm{i=b,s,r}{m_\textrm{i} \, r_\textrm{i}}) - (\sum_\textrm{i=b,s,r}{m_\textrm{i} \, k_\textrm{i}})\)
is independent of \(\phi\), and thus

\begin{equation}\label{eq:dPdt3}
\frac{\partial \left( \frac{ \textrm{d}P} {\textrm{d}t}\right)}{\partial \phi}  =
(B_4-1) a_\textrm{l} a_4\phi ^{-B_4}.
\end{equation}

Eq. \ref{eq:dPdt3} shows that if \(B4=1\), \(\frac{ \textrm{d}P} {\textrm{d}t}\) is independent
of $\phi$. In contrast, if \(B4>1\) (which tends to be the case empirically), \(\frac{ \textrm
{d}P} {\textrm{d}t}\) increases as $\phi$ increases, more acutely at low $\phi$.

Substituting into eq. \ref{eq:dg4}, we have the result that growth rate correlates
with traits if

\begin{equation} \label{eq:G6}
\frac{(B_4-1) a_\textrm{l} a_4\phi ^{-B_4}}{c_2 - a_\textrm{l} a_4 \phi ^{1-B_4}}
< \frac{1}{\phi
 + \frac{\textrm{d}m_\textrm{s}}{\textrm{d}a_\textrm{l}} + \frac{\textrm{d}m_\textrm
 {b}}{\textrm{d}a_\textrm{l}} + \frac{\textrm{d}m_\textrm{r}}{\textrm{d}a_\textrm{l}}}.
\end{equation}

\subsubsection{Wood construction cost}

Do similar to previous section for wood construction cost.
